Pandemic Pass? Treaty Derogations and Human Rights Practices During COVID-19

### 2023-03-08 manuscript/manuscript.qmd
Audrey and Suparna edits and additions

---
title: "Pandemic Pass? Treaty Derogations and Human Rights Practices During COVID-19"
subtitle: "Research note"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  email: schaudhry@lclark.edu
- name: Audrey L. Comstock
  affiliation: Arizona State University
  email: audrey.comstock@asu.edu
- name: Andrew Heiss
  affiliation: Georgia State University
  email: aheiss@gsu.edu
abstract: |
  In an effort to combat the COVID-19 pandemic, many countries declared states of emergency and derogated (temporarily suspended) from their international human rights treaty obligations. However, the abuse of emergency derogations by many governments raises concerns about the long-lasting impact of these sweeping measures on human rights. Does a robust civil society mitigate the adverse effects arising from international treaty derogations? Previous literature finds that an effective civil society response can mitigate the impact of natural disasters and crises and even increase political engagement in the long-run. Using data from the Varieties of Democracy PanDem dataset and the Oxford Covid-19 Government Response Tracker, we first assess the variation in human rights responses among governments during the pandemic. We then assess the role of civil society in keeping governments accountable and mitigating the effects of policies that violate human rights. Our results hold important implications for understanding government compliance with international treaties and shedding light on how states of emergency may directly and indirectly lead to democratic backsliding.
date: March 5, 2023
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/mountainous-mackerel>"
bibliography: bib.bib
link-citations: true
format: 
  html:
    appendix-style: plain
    toc: true
  pdf:
    mainfont: "IBM Plex Serif"
    sansfont: "Noto Sans"
    linestretch: 1.25
    toc: false
language: 
  title-block-published: "Latest version"
---

@@

Combating a deadly virus like SARS-CoV-2 requires extraordinary public health measures. In an effort to fight the COVID-19 pandemic, governments across the world imposed restrictions on the freedom of movement, association, assembly and privacy. While countries are ostensibly free to instate these policies within their own borders, the imposition of these restrictions raised important questions about states' commitments to international human rights principles. Some human rights treaties recognize states' needs for emergency measures when confronted with public health concerns and contain escape clauses that establish safeguards to prevent permanent erosion of these rights during emergencies. This temporary suspension of state compliance with its international treaty provisions occurs through a process known as derogations. However, the abuse of emergency measures during these derogations raises concerns about the long-lasting impact of these measures on human rights. Under emergency decrees, states often impose restrictions that are temporary in theory but difficult to roll back or contain in practice [@Stasavage2020; @LuhrmannRooney2021].

Given recent medical advancements and dwindling concerns about the pandemic, the Biden administration announced that in May 2023 it would be ending the emergency measures instated at the beginning of the pandemic—a move that that has occurred in many other countries as well. With the lifting of these measures, we can assess important descriptive and analytical questions raised regarding states' respect for international law and human rights during this period. What did derogations mean for compliance? Are derogating and non-derogating states different in terms of how they implemented emergency measures during COVID-19? Were derogating states more likely to exploit the provisions and to abuse human rights during COVID-19? These questions are important for the study of international law, compliance, and times of crises.

The violations of human rights during this period also raises questions about factors that can mitigate the adverse impacts of emergency measures. Existing literature on emergencies such as wars and natural disasters shows that civil society organizations (CSOs) play a pivotal role during these periods [@pellingdill; @cash2013]. They provide services and resources, and hold governments accountable to ensure inclusive policies during these emergencies. Can these groups also prevent or reduce the likelihood of human rights violations during public health emergencies? Does a robust civil society mitigate the adverse effects arising from international treaty derogations during the pandemic?

Typically, when international relations scholars write about human rights treaty compliance, they write of positive human rights recognition. Compliance is an adherence to the expectations and written standards within the treaties, often described as a "state of conformity or identity between an actor’s behavior and a specified rule" [@RaustialaSlaughter:2002, 539]. However, there are treaty qualifications that are legal addendums to commitment that alter or even limit compliance expectations. When states comply with these legal actions, they are in fact, limiting their obligations to human rights standards. In this research note, we focus on one such legal addendum—the treaty derogation. If and when states adhere to the written expectations of treaty derogations, they will be suspending human rights recognition during asset time of domestic emergency and crisis. If states are complying the derogation expectations, we should in fact expect a reduction in compliance with the overall treaty during that time and human rights standards. We explore how the potential decline in compliance may actually support human rights during the COVID-19 period since emergency mitigation strategies were targeting a reduction in disease outbreak. Human rights compliance literature generally finds that NGOs and civil society are key actors in facilitating treaty obligation actualization [e.g., @KeckSikkink:1998; @Merry:2009]. We explore what, if any, role civil society organizations had during this period.

Using data from the Varieties of Democracy PanDem dataset, the Oxford Covid-19 Government Response Tracker, and United Nations treaty derogation data we first assess the variation in human rights responses among governments during the pandemic. We then analyze the effect of derogating (or not) on human rights practices during the pandemic and assess the role of civil society in mitigating the effects of rights-violating policies. Our main takeaways are that derogating states did largely what they signaled they would through the derogations. States that derogated and specified time limits to the emergency provisions generally stuck to the time limits and did implement the emergency practices they signaled they would. Additionally, we find that, on average, derogating states are less likely to abuse emergency measures. On the other hand, states that did not derogate from their international human rights obligations and still impose emergency measures are more likely impose discriminatory policies, violate non-derogable rights (the right to life, freedom from torture, and freedom from slavery), and impose emergency measures without time limits. We also find that civil society can play an important mitigating role in curbing discriminatory policies. States with low levels of civil society repression are the least likely to abuse human rights when enforcing emergency policies. The likelihood of discriminatory policies increase as civil society repression worsens, but derogations appear to protect against worsening abuses—even in countries with severe civil society repression, those that derogate have a nearly 100% probability of no discriminatory policies.

This research note makes three contributions. First, while there is a significant literature on compliance with international law, especially human rights treaties [e.g. @Simmons2009; @Comstock2021], and the flexibility tools such as reservations, understandings, and declarations (RUDs) that states invoke when they join or withdraw from such treaties [@Neumayer:2007; @Helfer:2002; @Helfer:2005], we know little about states' compliance with flexibility mechanisms such as derogations. Compliance with derogations may not necessarily be the same as compliance with the treaty and this research note provides a systematic understanding of the types of states that are likely to comply with these escape clauses. The pandemic further provides unique insight into derogation behavior because numerous states derogated under the same global conditions of a public health emergency, rather than domestic conflict or natural disaster typical of treaty derogations, making it easier to compare the long-term human rights effects of these derogations cross nationally.

Second, there is a robust literature on compliance with international law and how it can lead to improvements in domestic human rights conditions [@Simmons2009; @Merry:2009]. But we know much less about how international treaty provisions intended to safeguard rights might actually worsen human rights outcomes. While derogations are intended to make sure government responses are temporary, proportional and non-discriminatory, some legal scholars warn of the possibility that some treaty actions removing legal obligations may actually contribute to more violations and abuse [e.g. @Lebret:2020; @helfer2021] as part of a broader trend of targeting groups during times of crises [@Grasse2021]. TODO: EXPAND ON RESULT In this scenario, when international factors may not be enough to constrain state actions, assessing the effect of domestic-level factors such as a robust civil society that potentially can mitigate the adverse effects of derogations becomes essential. TODO: WE FIND.

Finally, studying states' compliance with their international human rights treaty obligations as well as the (unintended) effects of derogations is important given trends in democratic backsliding and declining respect for human rights. Nearly 75 percent of the world’s population now lives in a country that faced democratic deterioration last year [@freedomhouse2021]. The pandemic led to further backsliding with many countries around the world postponing elections, arresting protestors, and cracking down on citizens to consolidate power—all under the guise of public health measures. These periods of emergencies may be an important piece to understand the processes of autocratization and democratic backsliding since leaders often use emergency decrees to  allow states to aggrandize executive power [@LuhrmannRooney2021; @LevitskyZiblatt:2018; @Bermeo2016]. The temporary suspension of states' international legal obligation is one possible pathway that could lead to long-term rights violations and democratic erosion.

Below, we first explore the purpose of derogations during the pandemic. In doing so, we focus specifically on derogations to the International Covenant on Civil and Political Rights (ICCPR)—arguably the most significant and extensive treaty in the international human rights regime. Following this, we introduce our expectations regarding how formally derogating from the treaty (or not) may impact a state's human rights practices during the pandemic. We then assess states' behavior towards civil society and these organizations' role in mitigating the adverse effects of emergency measures. We conclude with the implications of our findings for the study of international law, times of crises, and the role of civil society.

@@

Derogations in international law authorize states to temporarily suspend their international treaty commitments during times of crises, providing them with flexibility in responding to events. These times of crises include wars, natural disasters, internal political unrest such as terrorism, and public health emergencies. Derogations afford states the opportunity to adjust to these changing conditions [@HBHelferFariss]. Once states derogating from their international legal obligations, it also provides vital information to both international and domestic monitoring bodies, interest groups, and advocates about which rights are suspended, for how long, and the reasoning behind these suspensions. This information allows these actors—at least in principle—to challenge measures that are excessive, vague, or outlast the intended time frame of their implementation [@helfer2021].

While legal critical assessments of the use of derogations find importance in their use [@richardson2020], others argue that, "When a state does not bother with derogation provisions in times of emergencies where human rights are suspended, then it is a situation of unbridled state power.” [@burchill2005, 97] An abuse of state power can also potentially led to the violation of non-derogable rights—rights such as right to life, freedom from torture, freedom from slavery, freedom of thought and conscience, among others, that cannot be suspended even during emergencies. Research finds results that support this notion—there is no meaningful variation in states' respect for derogable or non-derogable rights during officially declared states of emergencies [@richardsclay2012; @neumayer2013]. Even states that officially derogate from their international legal obligations are more likely to violate physical integrity rights, which are non-derogable rights [@neumayer2013; @comstock2019].

These findings suggest that states may not be discerning the significance of derogable v/s. non-derogable rights status during emergencies. This also raises the question on the extent to which this distinction is a valuable one during the pandemic—which is a public health emergency different from other states of emerency normally brought on by war or internal political unrest. Overall, there is a general expectation that following derogations, we may observe some decline in respect for human rights.

@@

Research finds that democracies, especially those with strong rule of law, are more likely to derogate [@HBHelferFariss]. The expectation is that states with strong democratic institutions need to convince domestic audiences of the legality and legitimacy of suspending rights during emergencies.

A first question of interest before we examine the potential impact of derogations is what states are more likely to make them in the first place. Though not much research has focused on derogations in political science, most of it has focused on the determinants of submitting derogations to the UN rather than implications of their application. This literature tells us that states that are more democratic are more likely to issue derogations and other post-commitment actions such as reservations, understandings, and declarations [@Simmons2009; @comstock2019] and that most states facing an emergency do not, in fact, issue derogations [@HBHelferFariss]. The argument is that democratic states understand international law and international legal mechanisms to be more legitimate and recognize their importance more so than other states. Less democratic states, even though at times may be more likely to violate or evade international human rights law, are less likely to see nuanced pieces of treaty law as important. Looking to current derogation practices, there is not a perfect overlap in the states that issued derogations in the past towards the ICCPR and those that did during the COVID-19 pandemic. Of course, since derogations are a type of action submitted after treaty commitment, the only states that would be submitting them are ones that already have ratified or otherwise committed to the ICCPR. As of 2023, there are 173 states party to the ICCPR. This means that most, but not all of the UN’s 193 members had access to submitting derogations during COVID-19. Other core human rights treaties do not contain the same requirement to submit derogation notifications whenever an anticipated violation will occur.

Some states new to derogations introducing them during the COVID-19 pandemic [@Comstock:2023]. However, past treaty action behavior may help explain entering derogations during this time. To examine this first step, we code two variables 1) past ICCPR derogation behavior and 2) past ICCPR post-commitment action behavior. States receive a 1 if they engaged in this legal behavior before 2020 and 0 if they did not. This will allow us to empirically test for the first step of engagement before analyzing any impact COVID-19 era derogations had on different rights practices. There were 36 states that submitted ICCPR derogations and 60 that submitted any type of post-commitment action between 1966-2019.

@@

In this research note, we look at states' derogations from the International Covenant on Civil and Political Rights (ICCPR) during the pandemic. ICCPR is arguably one of the most extensive human rights treaties and the exceptional circumstances brought on by the global pandemic have led to numerous extensive derogations to the ICCPR. The ICCPR was adopted by the UN General Assembly on December 19, 1966 and came into force on March 23, 1976. It was designed to recognize "the inherent dignity and of the equal and inalienable rights of all members of the human family" and outlined dozens of articles aimed at creating conditions where everyone can enjoy their civil, political, economic, social, and cultural rights. The ICCPR is one of only a few select treaties that requires states to report on the suspension or derogation of certain civil and political rights only under specific situations of "public emergency which threatens the life of the nation" [@iccpr]. The other treaties with this requirement include the European Convention Protection of Human Rights and Fundamental Freedoms and the American Convention on Human Rights. However, the ICCPR is the only one with global, rather than regional, membership and participation to track during the pandemic.

To legally derogate, government must first officially declare a state of emergency (Article 4). These derogations must fulfill three criteria: they should be temporary in application, proportional to a crisis' actual threat and non-discriminatory. However, it is worth noting that states cannot derogate from the entirety of the ICCPR. Seven rights as enshrined in the ICCPR are non-derogable: right to life, prohibit torture or cruel, inhumane punishment, prohibit slavery, provide the right to recognition as a person before the law, and protect freedom freedom of thought, conscience and religion. On the other hand, several ICCPR Articles provide that public health needs can justify limitations on or derogations from certain rights [@richardson2020, 112]: freedom of movement (Article 12), freedom of expression (Article 19), and the rights to peaceful assembly and freedom of association (Article 22).

Recognizing that public health exigencies would require states to curb civil and political rights, the UN Human Rights Office of the High Commissioner (OHCHR) published guidance stating that any emergency measures must meet specific criteria of legality, necessity, proportionality, and non-discrimination. However,not every country formally derogated—a number of countries (HOW MANY) implemented emergency measures without formally derogating from ICCPR. The impact of these emergencies may have longer term impacts beyond temporary human rights violations [@BrownBrechenmacherCarothers:2020]. Below, we first theorize how derogations can protect against rights violations and the excesses of emergency measures, followed by the role of civil society in mitigating these violations.

@@

If states determine that emergency measures are justified without regard for international law, then we expect that states which did not file for derogations to the ICCPR were less likely to meet the criteria for derogations—necessity, proportionality, non-discrimination and time delimited, and therefore more likely to impose measures that violated these criteria.

@@

Given the divergent state responses during the pandemic, these differing responses can have lasting impacts on human rights trends. In the absence of adequate review from the international community during the pandemic, certain domestic-level factors may be necessary to mitigate the adverse effects of emergency measures. While both the theoretical and empirical literature has established that states may violate both derogable and non-derogable rights during emergencies, we have comparatively little evidence on factors that may safeguard against these rights violations during periods of derogation.[^exceptions]

There is a robust literature on how civil society is important for improving human rights outcomes [@KeckSikkink:1998; @RisseRoppSikkink; @landman2005; @HBTsutsui; @neumayer2005; @Jetschke2011]. Previous literature finds that an effective civil society response can mitigate the impact of natural disasters and crises and even increase political engagement in the long-run [@fair2017].

As such, we expect that a strong and vibrant civil society may help mitigate the adverse impact arising from derogations and emergency measures imposed by governments during the pandemic. Civil society organizations broaden the range of voices in the policymaking process and help amplify the concerns and interests of citizens [@murdie2014]. Civic initiatives have not only played a major role in countering the pandemic's severity, but have also pressured authorities into implementing better policies to deal with the impact of economic shutdowns, product shortages and spikes in unemployment. For instance, in Brazil, a broad coalition of CSOs lobbied the government to implement a relief fund, successfully securing a monthly stipend three times the amount initially suggested by the government [@carnegie2021, 5].

@@

We use three different primary independent variables to explain changes in these policy and human rights outcomes. For our preliminary models and in models related to our first hypothesis, we rely on a binary indicator of whether a state was under ICCPR derogation conditions during a given week. We collected and coded this data from real time state action updates recorded at the UN Treaty Collection [@UnitedNations:2023]. During the 15 months covered in this project, there were 115 formal derogations submitted mentioning 8 specific ICCPR articles; 15 derogations made no reference to any specific article. Of the derogations that indicated specific time limits, the average derogation duration was 41 days. We mark the start and end date of each derogation so that each country-week observation indicates whether a state has derogated.

@@

Finally, we include several other variables that are associated with and potentially confound both the decision to derogate and the decision to enact specific policies or human rights-related actions. To account for the weekly and overall severity of the pandemic in each country, we control for the count of new COVID cases and the count of new COVID deaths, as well as the cumulative count of cases and deaths. Since formally derogating from treaties can serve as a signal that countries take their obligations seriously, we include an indicator marking whether a country has ever derogated or taken other formal ICCPR bureaucratic actions prior to the pandemic. Finally, we include several annual measures of human rights, democracy, and civil society: V-Dem's rule of law index, civil liberties index, and core civil society index.

@@

In all model specifications, we include new and cumulative COVID cases and deaths as controls. In our first hypothesis, we add controls for previous ICCPR derogations or other actions, along with V-Dem indices for the respect of rule of law, civil liberties, and civil society. In our second hypothesis, we omit the core civil society index and instead include the interaction of civil society repression and derogations.

@@

The results of our models show that derogations have some protective effect against more bureaucratic forms of human rights abuses and less against practical abuses (see @fig-preds-mfx-h1). There is no substantial difference between derogating and non-derogating countries when considering the probability of imposing discriminatory policies or engaging in abusive enforcement. Both types of countries follow nearly identical trends over the first 15 months of the pandemic, with an 85% predicted probability of no discriminatory policy in March 2020 rising to nearly 100% by June 2021. The next most probable outcomes—minor and moderate levels of discriminatory policies—are equally likely regardless of derogation status. Moreover, countries that derogate were substantially less likely to engage in no abusive enforcement and instead turn to minor or moderate levels of abusive enforcement, particularly in the early months of the pandemic. Minor and moderate levels of excessive enforcement are each several percentage points more likely for derogating countries than for their non-derogating counterparts. The mechanisms behind this seeming reversal—where countries that follow correct human rights procedures end up enforcing emergency policies in a more abusive way—are unclear.

@@

# Discussion

>>>>

---
title: "Pandemic Pass? Treaty Derogations and Human Rights Practices During COVID-19"
subtitle: "Research note"
author:
- name: Suparna Chaudhry
  affiliation: Lewis and Clark College
  email: schaudhry@lclark.edu
- name: Audrey L. Comstock
  affiliation: Arizona State University
  email: audrey.comstock@asu.edu
- name: Andrew Heiss
  affiliation: Georgia State University
  email: aheiss@gsu.edu
abstract: |
  In an effort to combat the COVID-19 pandemic, many countries declared states of emergency and derogated (temporarily suspended) from their international human rights treaty obligations. Did states actually end up complying with their pandemic-era derogations? Did derogating and non-derogating states implement emergency measures differently, and if so, how?  Using data from the Varieties of Democracy PanDem dataset and the Oxford Covid-19 Government Response Tracker, we find that states which derogated from their international human rights obligations imposed emergency measures that were temporary and non-discriminatory. On the other hand, states that did not derogate were more likely impose discriminatory measures, enact emergency measures without time limits and violate non-derogable rights. Our results support the role that flexibility mechanisms such as derogations play in international law.
date: March 5, 2023
code-repo: "Access the code, data, and analysis at <https://github.com/andrewheiss/mountainous-mackerel>"
bibliography: bib.bib
link-citations: true
format: 
  html:
    appendix-style: plain
    toc: true
  pdf:
    mainfont: "IBM Plex Serif"
    sansfont: "Noto Sans"
    linestretch: 1.25
    toc: false
language: 
  title-block-published: "Latest version"
---

@@

# Introduction

Do emergency provisions during crises do more harm or good when it comes to human rights? Combating a deadly virus like SARS-CoV-2 required extraordinary public health measures. In an effort to fight the COVID-19 pandemic, governments across the world imposed restrictions on the freedom of movement, association, assembly and privacy. While countries were ostensibly free to instate these policies within their own borders, the imposition of these restrictions raised important questions about states' commitments to international human rights law. Some human rights treaties recognize states' needs for emergency measures during public health concerns and contain escape clauses that establish safeguards to prevent permanent erosion of these rights during emergencies. The temporary suspension of state compliance with its international treaty obligations occurs through a legal action known as derogation. However, the potential abuse of emergency measures during  derogations raises concerns about the long-lasting impact of emergency measures on human rights. Under emergency decrees, states often impose restrictions that are temporary in theory but difficult to roll back or contain in practice [@Stasavage2020; @LuhrmannRooney2021].

Given recent medical advancements and dwindling concerns about the pandemic, the Biden administration announced that it would end the public health emergency in the US on May 11, 2023, ending measures instituted at the beginning of the pandemic—a move that has occurred in many other countries as well. With the lifting of these measures, we can assess important descriptive and analytical questions raised regarding states' respect for international law and human rights during this emergency period. What did derogations mean for compliance? Did states deviate from treaty provisions when they said they would? Are derogating and non-derogating states different in terms of how they implemented emergency measures during COVID-19? Were derogating states more likely to exploit the provisions and to abuse human rights during COVID-19? These questions are important for the study of international law, compliance, and times of crises.

The potential violations of human rights during this period also raises questions about factors that can mitigate the adverse impacts of emergency measures. Existing literature on emergencies such as wars and natural disasters shows that civil society organizations (CSOs) play a pivotal role during these periods [@pellingdill; @cash2013]. They provide services and resources, and hold governments accountable to ensure inclusive policies during these emergencies. Human rights compliance literature also generally finds that NGOs and civil society are key actors in facilitating treaty obligation actualization [e.g., @KeckSikkink:1998; @Merry:2009]. Can these groups also prevent or reduce the likelihood of human rights violations during public health emergencies? Does a robust civil society mitigate the adverse effects arising from international treaty derogations during the pandemic?

Typically, when international relations scholars write about human rights treaty compliance, they write of positive human rights recognition. Compliance is an adherence to the expectations and written standards within the treaties, often described as a "state of conformity or identity between an actor’s behavior and a specified rule" [@RaustialaSlaughter:2002, 539]. However, there are treaty qualifications that are legal addendums to commitment that alter or even limit compliance expectations. When states comply with these legal actions, they are in fact, limiting their obligations to human rights standards. In this research note, we focus on one such legal addendum—the treaty derogation. If and when states adhere to the written expectations of treaty derogations, they will be suspending human rights recognition during a set time of domestic emergency and crisis. If states are complying with the derogation expectations, we should in fact expect a reduction in compliance with the overall human rights treaty during that time. We explore how the potential decline in compliance may actually support human rights during the COVID-19 period since emergency mitigation strategies also require certain rules and standards to prevent the abuse of rights.

Using data from the Varieties of Democracy PanDem dataset, the Oxford Covid-19 Government Response Tracker, and United Nations treaty derogation data, we first assess the variation in human rights responses among governments during the pandemic. We then analyze the effect of derogating (or not) on human rights practices during the pandemic and assess the role of civil society in mitigating the effects of rights-violating policies. We find that states that derogated and specified time limits to the emergency provisions generally stuck to the time limits and adhered to standards required during derogation by avoiding discriminatory policies, the abusive enforcement of these policies, and not violating non-derogable rights. On the other hand, states that did not derogate from their international human rights obligations and still imposed emergency measures were more likely impose discriminatory policies, violate non-derogable rights (the right to life, freedom from torture, and freedom from slavery), and impose emergency measures without time limits.

We also find that civil society can play an important mitigating role in curbing discriminatory policies. States with low levels of civil society repression were the least likely to abuse human rights when enforcing emergency policies. The likelihood of discriminatory policies increased as civil society repression worsened, but derogations appear to protect against worsening abuses—even in countries with severe civil society repression, those that derogated had a nearly 100% probability of no discriminatory policies.

This research note makes two contributions. First, while there is a significant literature on compliance with international law, especially human rights treaties [e.g. @Simmons2009; @Comstock2021], and the flexibility tools such as reservations, understandings, and declarations (RUDs) that states invoke when they join or withdraw from such treaties [@Neumayer:2007; @Helfer:2002; @Helfer:2005], we know little about states' compliance with flexibility mechanisms such as derogations. Compliance with derogations may not necessarily be the same as compliance with the treaty and this research note provides a systematic understanding of the types of states that are likely to comply with these escape clauses. The pandemic further provides unique insight into derogation behavior because numerous states derogated under the same global conditions of a public health emergency, rather than domestic conflict or natural disaster typical of treaty derogations, making it easier to compare the long-term human rights effects of these derogations cross nationally.

Second, there is a robust literature on compliance with international law and how it can lead to improvements in domestic human rights conditions [@Simmons2009; @Merry:2009]. While derogations are intended to make sure government responses are temporary, proportional and non-discriminatory, some legal scholars warn of the possibility that some treaty actions removing legal obligations may actually contribute to more violations and abuse [e.g. @Lebret:2020; @helfer2021] as part of a broader trend of targeting groups during times of crises [@Grasse2021]. We find that overall, derogating states do not increase abuse and violations. This finding speaks to international legal compliance scholarship and supports the idea that states are being sincere about their intentions and not, byinlarge, using actions like derogation to cover abusive behavior.

The rest of the paper proceeds as follows. First, we explore the purpose of derogations during the pandemic. In doing so, we focus specifically on derogations to the International Covenant on Civil and Political Rights (ICCPR)—arguably the most significant and extensive treaty in the international human rights regime. Following this, we introduce our expectations regarding how formally derogating from the treaty (or not) may impact a state's human rights practices during the pandemic. We then assess states' behavior towards civil society and these organizations' potential role in mitigating the adverse effects of emergency measures. We conclude with the implications of our findings for the study of international law, times of crises, and the role of civil society.

@@

Derogations in international law authorize states to temporarily suspend their international treaty commitments during times of crises, providing them with flexibility in responding to events. Times of crises can include wars, natural disasters, internal political unrest such as terrorism, and public health emergencies. Derogations afford states the opportunity to adjust to these changing conditions [@HBHelferFariss]. When states derogate from their international legal obligations,the derogation provides vital information to both international and domestic monitoring bodies, interest groups, and advocates about which rights are suspended, for how long, and the reasoning behind these suspensions. This information allows these actors—at least in principle—to challenge measures that are excessive, vague, or outlast the intended time frame of their implementation [@helfer2021].

While legal critical assessments of derogations find importance in their use [@richardson2020], others argue that, "When a state does not bother with derogation provisions in times of emergencies where human rights are suspended, then it is a situation of unbridled state power.” [@burchill2005, 97] An abuse of state power can also potentially lead to the violation of non-derogable rights—rights such as right to life, freedom from torture, freedom from slavery, freedom of thought and conscience, among others, that are viewed as essential rights that cannot be suspended even during emergencies. Research finds results that support this notion—there is no meaningful variation in states' respect for derogable or non-derogable rights during officially declared states of emergencies [@richardsclay2012; @neumayer2013]. Even states that officially derogate from their international legal obligations are more likely to violate physical integrity rights, which are non-derogable rights [@neumayer2013; @comstock2019].

These findings suggest that states may not be discerning the significance of derogable v/s. non-derogable rights status during emergencies. This also raises the question of the extent to which this distinction is a valuable one during the pandemic. Most derogations to the ICCPR have been for internal conflict or natural disaster - is a public health emergency different from other states of emergency? Overall, there is a general expectation that following derogations, we may observe some decline in respect for human rights. However, previous studies have not explored the relationship between derogations and public health emergencies.

@@

A first question of interest before we examine the potential impact of derogations is what states are more likely to submit them in the first place. Though not much research has focused on derogations in political science, and research that does exist has mostly focused on the determinants of submitting derogations to the UN rather than implications of when they are in effect. This literature tells us that states that are more democratic are more likely to issue derogations and other post-commitment actions such as reservations, understandings, and declarations [@Simmons2009; @comstock2019] and that most states facing an emergency do not, in fact, issue derogations [@HBHelferFariss]. The expectation is that states with strong democratic institutions need to convince domestic audiences of the legality and legitimacy of suspending rights during emergencies. Additionally, democratic states understand international law and international legal mechanisms to be more legitimate and recognize their importance more so than other states. Less democratic states, even though at times may be more likely to violate or evade international human rights law, are less likely to see nuanced pieces of treaty law as important.

@@

In this research note, we look at states' derogations from the International Covenant on Civil and Political Rights (ICCPR) during the pandemic. ICCPR is a foundational human rights treaty with broad human rights coverage. Exceptional circumstances brought on by the global pandemic have led to numerous, extensive derogations to the ICCPR removing state obligations to important rights areas of assembly, movement, and others. The ICCPR was adopted by the UN General Assembly on December 19, 1966 and entered into force on March 23, 1976. It was designed to recognize "the inherent dignity and of the equal and inalienable rights of all members of the human family" and outlined dozens of articles aimed at creating conditions where everyone can enjoy their civil, political, economic, social, and cultural rights. The ICCPR is one of only a few select treaties that requires states to report on the suspension or derogation of certain civil and political rights only under specific situations of "public emergency which threatens the life of the nation" [@iccpr]. The other treaties with this requirement include the European Convention Protection of Human Rights and Fundamental Freedoms and the American Convention on Human Rights. However, the ICCPR is the only one with global, rather than regional, membership and participation to track during the pandemic. As of 2023, there are 173 states party to the ICCPR. This means that most, but not all of the UN’s 193 members had the ability to submit derogations during COVID-19. Other core human rights treaties do not contain the same requirement to submit derogation notifications whenever an anticipated violation will occur.

To legally derogate, government must first officially declare a state of emergency (Article 4). Derogations must fulfill three criteria: they should be temporary in application, proportional to a crisis' actual threat and non-discriminatory. However, it is worth noting that states cannot derogate from the entirety of the ICCPR. Seven rights enshrined in the ICCPR are non-derogable: the right to life, the prohibition of torture or cruel, inhumane punishment,the prohibition of slavery, the right to recognition as a person before the law, and the protection of freedom of thought, conscience and religion. The international legal community viewed these rights as so fundamental and important that no type of crisis could justify their suspension. Other types of rights are seen as able to be suspended during crises. Several ICCPR Articles provide that public health needs can justify limitations on or derogations from certain rights [@richardson2020, 112]: freedom of movement (Article 12), freedom of expression (Article 19), and the rights to peaceful assembly and freedom of association (Article 22).

Recognizing that public health exigencies would require states to curb civil and political rights, the UN Human Rights Office of the High Commissioner (OHCHR) published guidance stating that any emergency measures must meet specific criteria of legality, necessity, proportionality, and non-discrimination. However,not every country formally derogated—a number of countries implemented emergency measures without formally derogating from ICCPR. Over 100 countries issued emergency declarations while fewer than 30 submitted COVID-19 related derogations [@icnl-covid]. These COVID-19 emergency declarations may have longer term impacts beyond temporary human rights violations [@BrownBrechenmacherCarothers:2020]. Below, we first theorize how derogations can protect against rights violations and the excesses of emergency measures, followed by the role of civil society in mitigating these violations.

@@

We expect that states that did not notify the ICCPR of impending rights suspensions and submit derogations were less likely to take the the criteria for derogations—necessity, proportionality, non-discrimination and time delimited seriously, and therefore more likely to impose measures that violated these criteria. If states determined that emergency measures are justified without consideration of international law, we expect that there will be more violations.

@@

Given the divergent state responses during the pandemic, differing responses can have lasting impacts on human rights trends. In the absence of adequate review from the international community during the pandemic, certain domestic-level factors may be necessary to mitigate the adverse effects of emergency measures. While both the theoretical and empirical literature has established that states may violate both derogable and non-derogable rights during emergencies, we have comparatively little evidence regarding factors that may safeguard against these rights violations during periods of derogation.

There is a robust literature on how civil society is important for improving human rights outcomes [@KeckSikkink:1998; @RisseRoppSikkink; @landman2005; @HBTsutsui; @neumayer2005; @Jetschke2011]. Previous literature finds that an effective civil society response can mitigate the impact of natural disasters and crises and even increase political engagement in the long-run [@fair2017]. As such, we expect that a strong and vibrant civil society may help mitigate the adverse impact arising from derogations and emergency measures imposed by governments during the pandemic. Civil society organizations broaden the range of voices in the policymaking process and help amplify the concerns and interests of citizens [@murdie2014]. Civic initiatives have not only played a major role in countering the pandemic's severity, but have also pressured authorities into implementing better policies to deal with the impact of economic shutdowns, product shortages and spikes in unemployment. For instance, in Brazil, a broad coalition of CSOs lobbied the government to implement a relief fund, successfully securing a monthly stipend three times the amount initially suggested by the government [@carnegie2021, 5].

@@

We use three different primary independent variables to explain changes in policy and human rights outcomes. For our preliminary models and in models related to our first hypothesis, we rely on a binary indicator of whether a state was under ICCPR derogation conditions during a given week. We collected and coded this data from real time state action updates recorded at the UN Treaty Collection [@UnitedNations:2023]. During the 15 months covered in this project, there were 115 formal derogations submitted mentioning 8 specific ICCPR articles; 15 derogations made no reference to any specific article. Of the derogations that indicated specific time limits, the average derogation duration was 41 days. We mark the start and end date of each derogation so that each country-week observation indicates whether a state has derogated.

@@

Finally, we include several other variables that are associated with and potentially confound both the decision to derogate and the decision to enact specific policies or human rights-related actions. To account for the weekly and overall severity of the pandemic in each country, we control for the count of new COVID-19 cases and the count of new COVID-19 deaths, as well as the cumulative count of cases and deaths.

Since formally derogating from treaties can serve as a signal that countries take their obligations seriously, we include an indicator marking whether a country has ever derogated or taken other formal ICCPR post-commitment actions, such as RUDs, prior to the pandemic. Some states new to derogations introduced them during the COVID-19 pandemic [@Comstock:2023]. However, past treaty action behavior may help explain entering derogations during this time. To examine this first step, we code two binary variables: (1) past ICCPR derogation behavior before 2020 and (2) past ICCPR post-commitment action behavior before 2020. This allows us to empirically test for the first step of engagement before analyzing any impact COVID-19 era derogations had on different rights practices. There were 36 states that submitted ICCPR derogations and 60 that submitted any type of post-commitment action between 1966-2019. Finally, we include several annual measures of human rights, democracy, and civil society: V-Dem's rule of law index, civil liberties index, and core civil society index.

@@

In all model specifications, we include new and cumulative COVID-19 cases and deaths as controls. In our first hypothesis, we add controls for previous ICCPR derogations or other actions, along with V-Dem indices for the respect of rule of law, civil liberties, and civil society. In our second hypothesis, we omit the core civil society index and instead include the interaction of civil society repression and derogations.

@@

The results of our models show that derogations have some protective effect against more bureaucratic forms of human rights abuses and less against practical abuses (see @fig-preds-mfx-h1). There is no substantial difference between derogating and non-derogating countries when considering the probability of imposing discriminatory policies or engaging in abusive enforcement. Both types of countries follow nearly identical trends over the first 15 months of the pandemic, with an 85% predicted probability of no discriminatory policy in March 2020 rising to nearly 100% by June 2021. The next most probable outcomes—minor and moderate levels of discriminatory policies—are equally likely regardless of derogation status. Moreover, countries that derogate were substantially less likely to engage in no abusive enforcement and instead turn to minor or moderate levels of abusive enforcement, particularly in the early months of the pandemic. Minor and moderate levels of excessive enforcement are each several percentage points more likely for derogating countries than for their non-derogating counterparts.

The mechanisms behind this seeming reversal—where countries that follow correct human rights procedures end up enforcing emergency policies in a more abusive way may be attributable to principal-agent problems. While elites or diplomats are the ones submitting derogations and committing to complying with international law, they are not the actors implementing emergency measures. During the pandemic, there may be immense variation in states' abilities to enforce these policies as intended, especially when it relates to human rights. Lower-level officials such as those in the police, state security forces and regional bureaucrats are often the ones enforcing emergency measures. Unlike their principals (elites), these agents neither have training in international law nor incentives to comply with derogations. Further, as previous research shows, during high levels of stress and fear, decisionmaking is more likely to be driven by agents who discounts principals' (dis)incentives [@HooverGreen:2016]. As such, these differences in incentives and training between elites and local officials may explain the incidences of abusive enforcement even when states submit derogations and claim that emergency measures will be implemented proportional to the threat at hand.

@@

# Discussion and conclusion
